R for Clinical Study Reports and Submission

### 2023-05-22 project-folder.qmd
Migrate from bookdown to Quarto book (working HTML book)

# Project folder {#folder}

@@

## Consistency {#consistency}

@@

::: {.rmdnote data-latex=""}
Under the `vignettes/` folder, there are two folders: `adam/` and `tlf/`.
The `adam/` folder contains ADaM datasets.
The `tlf/` folder contains output TLFs in RTF format.
We put `adam/` and `tlf/` folders within the `vignettes/` folder only for illustration purposes.
In an actual A&R report, you may have a different location to save your input and output.
:::

@@

## Reproducibility {#reproduce}

@@

::: {.rmdnote data-latex=""}
This is the same level of reproducibility in most SAS environments:
<https://support.sas.com/en/technical-support/services-policies.html#altos>
:::

@@

::: {.rmdnote data-latex=""}
`.Rprofile` is only for project-specific R packages.
A standard R package should not use `.Rprofile`.
:::

@@

::: {.rmdnote data-latex=""}
RStudio Package Manager (RSPM) provides a solution to host both
publicly available and internally developed R packages.
However, the [public RSPM server](https://packagemanager.rstudio.com/client/#/repos/2/overview)
does not provide a daily snapshot as MRAN does.
:::

@@

::: {.rmdnote data-latex=""}
The `renv.lock` file and `renv/` folder are only for project-specific R package.
A standard R package should not use renv.
:::

@@

::: {.rmdnote data-latex=""}
A cloud-based R environment (e.g., RStudio Workbench)
can enhance the reproducibility within an organization
by using the same operating system, R version, and R package versions for an A&R project.
More details can be found at <https://environments.rstudio.com/>.
:::

::: {.rmdnote data-latex=""}
A container solution like Docker [@RJ-2020-007] could further enhance
the reproducibility across an organization at the operating system level
but beyond the scope of this book.
:::

>>>>

# Project folder {#sec-folder}

@@

## Consistency {#sec-consistency}

@@

::: {.callout-note}
Under the `vignettes/` folder, there are two folders: `adam/` and `tlf/`.
The `adam/` folder contains ADaM datasets.
The `tlf/` folder contains output TLFs in RTF format.
We put `adam/` and `tlf/` folders within the `vignettes/` folder only for illustration purposes.
In an actual A&R report, you may have a different location to save your input and output.
:::

@@

## Reproducibility {#sec-reproduce}

@@

::: {.callout-tip}
This is the same level of reproducibility in most SAS environments:
<https://support.sas.com/en/technical-support/services-policies.html#altos>
:::

@@

::: {.callout-caution}
`.Rprofile` is only for project-specific R packages.
A standard R package should not use `.Rprofile`.
:::

@@

::: {.callout-note}
Posit Package Manager provides a solution to host both publicly available
and internally developed R packages. However, the
[Posit Public Package Manager](https://packagemanager.posit.co/)
does not provide a daily snapshot as MRAN does.
:::

@@

::: {.callout-caution}
The `renv.lock` file and `renv/` folder are only for project-specific R package.
A standard R package should not use renv.
:::

@@

::: {.callout-tip}
A cloud-based R environment (e.g., RStudio Workbench)
can enhance the reproducibility within an organization
by using the same operating system, R version, and R package versions for an A&R project.
More details can be found at <https://environments.rstudio.com/>.
:::

::: {.callout-note}
A container solution like Docker [@RJ-2020-007] could further enhance
the reproducibility across an organization at the operating system level
but beyond the scope of this book.
:::

### 2023-05-22 project-management.qmd
Migrate from bookdown to Quarto book (working HTML book)

# Project management {#manage}

@@

The project lead should also set up a project folder, as discussed in Chapter \@ref(folder).
The project initiation can be simplified by creating an
[RStudio project template](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html).

To enable reproducibility, the project leader should also review the startup file
(i.e. `.Rprofile` discussed in Section \@ref(reproduce)) and define:

@@

::: {.rmdnote data-latex=""}
After project initiation, modifying `.Rprofile` will be a risk of reproducibility
and should be handled carefully if necessary.
:::

@@

If there are any project-specific R functions that need to be developed,
the R functions can be placed in the `R/` folder as discussed in Section \@ref(consistency).

>>>>

# Project management {#sec-manage}

@@

The project lead should also set up a project folder, as discussed in @sec-folder.
The project initiation can be simplified by creating an
[RStudio project template](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html).

To enable reproducibility, the project leader should also review the startup file
(i.e. `.Rprofile` discussed in @sec-reproduce) and define:

@@

::: {.callout-caution}
After project initiation, modifying `.Rprofile` will be a risk for
reproducibility and should be handled carefully if necessary.
:::

@@

If there are any project-specific R functions that need to be developed,
the R functions can be placed in the `R/` folder as discussed in
@sec-consistency.

### 2023-05-22 project-overview.qmd
Migrate from bookdown to Quarto book (working HTML book)

# (PART) Clinical trial project {-}

@@

Chapter \@ref(folder) will discuss how to organize source code, documents, and
deliverables in an A&R clinical project.
We recommend using the R package folder structure.

Chapter \@ref(manage) will discuss a process or system development lifecycle
to manage the A&R of a clinical project.
We recommend following an agile management approach to define, develop, validate, and deliver work.

>>>>

@@

@sec-folder will discuss how to organize source code, documents, and
deliverables in an A&R clinical project.
We recommend using the R package folder structure.

@sec-manage will discuss a process or system development lifecycle
to manage the A&R of a clinical project.
We recommend following an agile management approach to define, develop, validate, and deliver work.

### 2023-05-22 submission-environment.qmd
Migrate from bookdown to Quarto book (working HTML book)

# Running environment {#running-environment}

@@

`cleanslate` is an R package that offers a solution to
create portable R environments.

::: {.rmdnote data-latex=""}
As of Q4 2021, the `cleanslate` package used in this chapter
is still under active development and validation.
This chapter gives a preview of the planned APIs.
They may change in the future.
:::

Install `cleanslate` from CRAN (once available):

@@

The `cleanslate` package supports:

@@

An essential feature of `cleanslate` is that it does **not** require
administrator privileges to run R and Rtools installers.
This makes it easier to deploy under enterprise settings
and avoids security and portability concerns.

@@

If you use `cleanslate`, remember to update the ADRG instructions for
executing the analysis programs in R. Mostly, this can simplify
the first three steps on creating a project, installing a specific
version of R, and configuring the package repo location.
For example:

@@

To make it convenient to use `cleanslate` in experiments,
one can also use its RStudio addin. After `cleanslate`
is installed, click `Addins` -> `cleanslate` -> `Create portable R environment`
in RStudio, or call `cleanslate:::create_env_addin()` to open it.

>>>>

# Running environment {#sec-running-environment}

@@

cleanslate is an R package that offers a solution to
create portable R environments.

::: {.rmdnote data-latex=""}
As of Q4 2021, the cleanslate package used in this chapter
is still under active development and validation.
This chapter gives a preview of the planned APIs.
They may change in the future.
:::

Install cleanslate from CRAN (once available):

@@

The cleanslate package supports:

@@

An essential feature of cleanslate is that it does **not** require
administrator privileges to run R and Rtools installers.
This makes it easier to deploy under enterprise settings
and avoids security and portability concerns.

@@

If you use cleanslate, remember to update the ADRG instructions for
executing the analysis programs in R. Mostly, this can simplify
the first three steps on creating a project, installing a specific
version of R, and configuring the package repo location.
For example:

@@

To make it convenient to use cleanslate in experiments,
one can also use its RStudio addin. After cleanslate
is installed, click `Addins` -> `cleanslate` -> `Create portable R environment`
in RStudio, or call `cleanslate:::create_env_addin()` to open it.

### 2023-05-22 submission-overview.qmd
Migrate from bookdown to Quarto book (working HTML book)

# (PART) eCTD submission {-}

@@

::: {.rmdnote data-latex=""}
"Sponsors should provide the software programs used to create all ADaM
datasets and **generate tables and figures associated with primary and
secondary efficacy analyses**. Furthermore, sponsors should submit software
programs used to generate additional information included in Section 14
CLINICAL STUDIES of the Prescribing Information (PI)26 if applicable.
**The specific software utilized should be specified in the ADRG**.
The main purpose of requesting the submission of these programs
**is to understand the process by which the variables for the respective
analyses were created and to confirm the analysis algorithms**.
Sponsors should submit software programs in **ASCII text format**;
however, executable file extensions should not be used."
:::

Chapter \@ref(submission-package) will focus on preparing
proprietary R packages and analysis code into proper
formats for submission.

Chapter \@ref(running-environment) will discuss
the recommendations to make the R code running environment
reproducible for dry run tests and reviews.

>>>>

@@

> Sponsors should provide the software programs used to create all ADaM
> datasets and **generate tables and figures associated with primary and
> secondary efficacy analyses**. Furthermore, sponsors should submit software
> programs used to generate additional information included in Section 14
> CLINICAL STUDIES of the Prescribing Information (PI)26 if applicable.
> **The specific software utilized should be specified in the ADRG**.
> The main purpose of requesting the submission of these programs
> **is to understand the process by which the variables for the respective
> analyses were created and to confirm the analysis algorithms**.
> Sponsors should submit software programs in **ASCII text format**;
> however, executable file extensions should not be used.

@sec-submission-package will focus on preparing
proprietary R packages and analysis code into proper
formats for submission.

@sec-running-environment will discuss
the recommendations to make the R code running environment
reproducible for dry run tests and reviews.

### 2023-05-22 submission-package.qmd
Migrate from bookdown to Quarto book (working HTML book)

# Submission package {#submission-package}

@@

This chapter uses `pkglite` to convert R source packages
into text files and back.

@@

For dependencies hosted in private Git repositories, you should pack them
with the primary R package(s) you want to submit,
as `pkglite` supports packing multiple R packages into a single text file;
then restore and install them in the order they are packed.

@@

Always use the same snapshot package repo for developing the TLFs and for submission.
Again, this can be checked in the project startup script,
as discussed in section \@ref(reproduce).

>>>>

# Submission package {#sec-submission-package}

@@

This chapter uses pkglite [@zhao2023electronic] to convert R source packages
into text files and back.

@@

For dependencies hosted in private Git repositories, you should pack them
with the primary R package(s) you want to submit,
as pkglite supports packing multiple R packages into a single text file;
then restore and install them in the order they are packed.

@@

Always use the same snapshot package repo for developing the TLFs and for submission.
Again, this can be checked in the project startup script,
as discussed in @sec-reproduce.

### 2023-05-22 tlf-ae-specific.qmd
Migrate from bookdown to Quarto book (working HTML book)

@@

For illustration purpose, we only provide counts in the simplified
table. The percentage of participants for each AE can be
calculated as shown in Chapter \@ref(aesummary).

@@

```{r}
ana <- adae %>%
  mutate(
    AESOC = toTitleCase(tolower(AESOC)),
    AEDECOD = toTitleCase(tolower(AEDECOD))
  )

@@

>>>>

```{r, include=FALSE}
source("_common.R")
```

@@

For illustration purpose, we only provide counts in the simplified
table. The percentage of participants for each AE can be
calculated as shown in @sec-aesummary.

@@

```{r}
fmt_num <- function(x, digits, width = digits + 4) {
  formatC(
    x,
    digits = digits,
    format = "f",
    width = width
  )
}
```

```{r}
ana <- adae %>%
  mutate(
    AESOC = tools::toTitleCase(tolower(AESOC)),
    AEDECOD = tools::toTitleCase(tolower(AEDECOD))
  )

@@

```{r}
count_by <- function(data, # Input data set
                     grp, # Group variable
                     var, # Analysis variable
                     var_label = var, # Analysis variable label
                     id = "USUBJID") { # Subject ID variable
  data <- data %>% rename(grp = !!grp, var = !!var, id = !!id)

  left_join(
    count(data, grp, var),
    count(data, grp, name = "tot"),
    by = "grp",
  ) %>%
    mutate(
      pct = fmt_num(100 * n / tot, digits = 1),
      n = fmt_num(n, digits = 0),
      npct = paste0(n, " (", pct, ")")
    ) %>%
    pivot_wider(
      id_cols = var,
      names_from = grp,
      values_from = c(n, pct, npct),
      values_fill = list(n = "0", pct = fmt_num(0, digits = 0))
    ) %>%
    mutate(var_label = var_label)
}
```

### 2023-05-22 tlf-assemble.qmd
Migrate from bookdown to Quarto book (working HTML book)

```{r, include=FALSE}
source("common.R")
```

@@

::: {.rmdnote data-latex=""}
The code below requires r2rtf version >= 1.0.0.
:::

@@

::: {.rmdnote data-latex=""}
One limitation of combining RTF source code is that
we are not able to specify the page orientation of each TLF
in the combined document.
:::

@@

::: {.rmdnote data-latex=""}
There is a slight learning curve on how toggle fields work
in Microsoft Word.
After you become familiar with the workflow,
toggle fields can extend your capability to manage a large number of TLFs
in RTF format.
:::

@@

::: {.rmdnote data-latex=""}
A typical error message is that system can not find the file
if you only provide a relative path.
Please double-check that the correct absolute file path is in the `INCLUDETEXT` field.
:::

@@

::: {.rmdnote data-latex=""}
If you modify table content in the combined `.docx` file,
you may get a weird table layout if you update all fields within a toggle field.
To resolve the issue, please remove all `\* MERGEFORMAT` in the filed mode using Alt + F9 before updating all toggle fields.
:::

>>>>

```{r, include=FALSE}
source("_common.R")
```

@@

::: {.callout-note}
The code below requires r2rtf version >= 1.0.0.
:::

@@

::: {.callout-caution}
One limitation of combining RTF source code is that
we are not able to specify the page orientation of each TLF
in the combined document.
:::

@@

::: {.callout-tip}
There is a slight learning curve on how toggle fields work
in Microsoft Word.
After you become familiar with the workflow,
toggle fields can extend your capability to manage a large number of TLFs
in RTF format.
:::

@@

::: {.callout-tip}
A typical error message is that system can not find the file
if you only provide a relative path.
Please double-check that the correct absolute file path is in the `INCLUDETEXT` field.
:::

@@

::: {.callout-tip}
If you modify table content in the combined `.docx` file,
you may get a weird table layout if you update all fields within a toggle field.
To resolve the issue, please remove all `\* MERGEFORMAT` in the filed mode using Alt + F9 before updating all toggle fields.
:::

### 2023-05-22 tlf-efficacy-ancova.qmd
Migrate from bookdown to Quarto book (working HTML book)

To prepare the report, we create a few helper functions
by using the `fmt_num()` function defined in Chapter \@ref(population).

@@

@@

```{r}
t2 <- fit_between %>%
  as_tibble() %>%
  mutate(
    ls = fmt_ci(
      estimate,
      estimate - 1.96 * SE,
      estimate + 1.96 * SE
    ),
    p = fmt_pval(p.value)
  ) %>%
  filter(str_detect(contrast, "- Placebo")) %>%
  select(contrast, ls, p)

@@

```{r}
t1_rtf <- t1 %>%
  data.frame() %>%
  rtf_title(c(
    "ANCOVA of Change from Baseline Glucose (mmol/L) at Week 24",
    "LOCF",
    "Efficacy Analysis Population"
  )) %>%
  rtf_colheader("| Baseline | Week 24 | Change from Baseline",
    col_rel_width = c(2.5, 2, 2, 4)
  ) %>%
  rtf_colheader(paste(
    "Treatment |",
    paste0(rep("N | Mean (SD) | ", 3), collapse = ""),
    "LS Mean (95% CI){^a}"
  ),
  col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)
  ) %>%
  rtf_body(
    text_justification = c("l", rep("c", 7)),
    col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)
  ) %>%
  rtf_footnote(c(
    "{^a}Based on an ANCOVA model after adjusting baseline value. LOCF approach is used to impute missing values.",
    "ANCOVA = Analysis of Covariance, LOCF = Last Observation Carried Forward",
    "CI = Confidence Interval, LS = Least Squares, SD = Standard Deviation"
  ))

>>>>

To prepare the report, we create a few helper functions
by using the `fmt_num()` function defined in @sec-population.

@@

```{r}
fmt_num <- function(x, digits, width = digits + 4) {
  formatC(
    x,
    digits = digits,
    format = "f",
    width = width
  )
}
```

@@

```{r}
t2 <- fit_between %>%
  as_tibble() %>%
  mutate(
    ls = fmt_ci(
      estimate,
      estimate - 1.96 * SE,
      estimate + 1.96 * SE
    ),
    p = fmt_pval(p.value)
  ) %>%
  filter(stringr::str_detect(contrast, "- Placebo")) %>%
  select(contrast, ls, p)

@@

```{r}
t1_rtf <- t1 %>%
  data.frame() %>%
  rtf_title(c(
    "ANCOVA of Change from Baseline Glucose (mmol/L) at Week 24",
    "LOCF",
    "Efficacy Analysis Population"
  )) %>%
  rtf_colheader("| Baseline | Week 24 | Change from Baseline",
    col_rel_width = c(2.5, 2, 2, 4)
  ) %>%
  rtf_colheader(
    paste(
      "Treatment |",
      paste0(rep("N | Mean (SD) | ", 3), collapse = ""),
      "LS Mean (95% CI){^a}"
    ),
    col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)
  ) %>%
  rtf_body(
    text_justification = c("l", rep("c", 7)),
    col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)
  ) %>%
  rtf_footnote(c(
    "{^a}Based on an ANCOVA model after adjusting baseline value. LOCF approach is used to impute missing values.",
    "ANCOVA = Analysis of Covariance, LOCF = Last Observation Carried Forward",
    "CI = Confidence Interval, LS = Least Squares, SD = Standard Deviation"
  ))

### 2023-05-22 tlf-overview.qmd
Migrate from bookdown to Quarto book (working HTML book)

\mainmatter

# (PART) Delivering TLFs in CSR {-}

@@

::: {.rmdnote data-latex=""}
FDA's PDF specification is a general reference. Each organization can define
more specific TLF format requirements that can be different from the examples in this book.
:::

@@

::: {.rmdnote data-latex=""}
There are other R packages to create TLFs in ASCII, RTF and Word format.
For example, rtables, huxtable, pharmaRTF, gt, officer, flextable etc.
Here we focus on r2rtf to illustrate the concept.  
Readers are encouraged to explore other R packages to find the proper tools to fit your purpose. 
:::

>>>>

@@

::: {.callout-note}
FDA's PDF specification is a general reference. Each organization can define
more specific TLF format requirements that can be different from the examples in this book.
:::

@@

::: {.callout-note}
There are other R packages to create TLFs in ASCII, RTF and Word format.
For example, rtables, huxtable, pharmaRTF, gt, officer, flextable etc.
Here we focus on r2rtf to illustrate the concept.
Readers are encouraged to explore other R packages to find the proper tools to fit your purpose.
:::
